Ricci-Ollivier Curvature of two Random Walks on the Rooted Phylogenetic Subtree-Prune-Regraft Graph

### 2015-03-28 curvature.tex
A full draft! (I think)

\usepackage{amsmath,amsthm}
\usepackage{csvsimple}
\usepackage{ifthen}
\usepackage{todonotes}
%\usepackage[notref,notcite]{showkeys} % show keys for eqs, etc.
\usepackage[right]{showlabels}
\usepackage{cite}
\usepackage{enumitem}
\usepackage{url}
\usepackage{subfigure}
\usepackage[nomarkers,nofiglist]{endfloat}
\let\MYoriglatexcaption\caption
\renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
% shift for todonotes
\usepackage[left=0.5in,right=1.5in,top=1in,bottom=1in,marginparwidth=1.25in]{geometry}
% regular margins
%\usepackage[margin=1in]{geometry}

@@

% % Make a todo table of contents.
% \makeatletter
% \providecommand\@dotsep{5}
% \makeatother
% \listoftodos\relax

@@

Ollivier-Ricci curvature provides an intuitive formalization for quantifying the shape of a graph with respect to a random walk (or indeed any metric space, see~\cite{Ollivier2009-bw}).
Let $m_x$ and $m_y$ be probability densities of a specified random walk starting at points $x$ and $y$ of a graph $G = (V,E)$, respectively.
The earth movers distance~\cite{rubner2000earth} (equivalently $L^1$ transportation, or Wasserstein distance~\cite{wasserstein}) between $m_x$ and $m_y$ is the minimum amount of ``work'' required to move $m_x$ to $m_y$ following the graph, that is
 $$ W_1(m_x, m_y) := \min_{\xi \in \Pi(m_x, m_y)} \sum \dspr{x,y} \xi(x,y),$$
where $\Pi(m_x, m_y)$ is the set of of assignments of probability from $m_x$ to pairs of vertices in $V \times V$.

@@

\subsection{Access time computation}
The access time for a pair of vertices in a graph is the (random) number of iterations required to go from one of the vertices to the other in a random walk~\cite{lovasz1993random}; we were interested in the connection between curvature and access time.
In previous work, we computed mean access times (MAT) between pairs of trees in the MCMC random walks---the mean number of iterations required to move from one tree to the other.
We applied this work to demonstrate graph effects in uSPR subgraphs of real MCMC samples using MrBayes~\cite{Whidden2015-yi} using the \texttt{sprspace}~\cite{sprspace} software.

Here, to gain more insight, we used simulation to approximate the entire access time distribution.
As before, calculating these distributions for all pairs of trees would have required a tremendous amount of memory and computational power in order to get good estimates.
Here again we use the insight that the access time for a pair of trees with a simple random walk does not depend on the actual labeling of those trees, but rather only on their relative labeling.
Thus rather than enumerate access times between trees, we enumerate times between pairs of trees in a tanglegram.
For example, because the pairs
$\{(((1,2),3),4), ((1,2),(3,4))\}$ and
$\{(((1,4),3),4), ((1,4),(2,3))\}$
are the same after relabeling, their access times will have identical distributions.
Thus to calculate the empirical distributions of access times we aggregate all access times for the same tangle \todo{cite software}.

@@

\subsection{Results}

Now

\begin{figure}
	\hspace*{\stretch{1}}
	\subfigure[\label{fig:short-time-kappa-access}]{\includegraphics[width=3in]{figs/short-time-kappa-access}}
	\hspace*{\stretch{2}}
	\subfigure[\label{fig:long-time-kappa-access}]{\includegraphics[width=3in]{figs/long-time-kappa-access}}
	\hspace*{\stretch{1}}
    \caption{Access times for degree 24 trees and $\kappa$.}
	\label{fig:kappa-access}
\end{figure}

\begin{table}
\centering
\caption{Regression showing factors determining mean access time for rSPR.}
\subtable[5 taxa]{\csvautotabular{analysis/5-walks-200K/rspr/mean-access-5-regression.csv}}
\subtable[6 taxa]{\csvautotabular{analysis/6-walks-50K/rspr/mean-access-6-regression.csv}}
\label{TABregressionMAT}
\end{table}

\begin{table}
\centering
\caption{Regression showing factors determining $\delta_1$ for rSPR.}
\subtable[5 taxa]{\csvautotabular{analysis/5-walks-200K/rspr/delta1-5-regression.csv}}
\subtable[6 taxa]{\csvautotabular{analysis/6-walks-50K/rspr/delta1-6-regression.csv}}
\label{TABregressionDelta}
\end{table}

\section{Rooted SPR Neighborhoods}

Having explored curvature values empirically for SPR graphs with 3--7 leaves, we now consider them theoretically.
In the absence of a bias provided by a likelihood function, curvature for the random walks we consider is entirely determined by the degree of vertices in the graph.
Thus, in order to provide bounds on curvature, we will begin by bounding differences between degrees.

@@

	We obtain the upper bounds similarly to the lower bounds, by observing that the affected at most $\frac{1}{6}$ of the mass may move through at most two fewer trees (i.e. directly between $T$ and $S$ rather than a pair of neighbors at distance $\dspr{T,S} + 2$ from each other).
	Again, this is at most one fewer tree when $\dspr{T,S}=1$.

\subsection{Computational investigation}
We computed all of the curvatures for pairs of trees with up to seven leaves to confirm our theoretical work as described before. \todo{Update once 7 taxon runs have finished}.

@@

\begin{figure}
	\hspace*{\stretch{1}}
	\subfigure[5 taxa]{\includegraphics[width=3in]{figs/nni-upmh-scatter5}}
	\hspace*{\stretch{2}}
	\subfigure[6 taxa]{\includegraphics[width=3in]{figs/nni-upmh-scatter6}}
	\hspace*{\stretch{1}}
    \caption{Scatterplot of $\kappa(\MH; T_1,T_2)$ values versus $\dnni{T_1,T_2}$ for the NNI graph. Distance values randomly perturbed (``jittered'') a small amount to avoid overplotting.}
	\label{fig:nni-scatter}
\end{figure}

\section{Conclusion and future work}
This isn't quite the right graph to be looking at (rooted).

Want to investigate curvature of MCMC steps with a nontrivial likelihood function.
MCMC does both branch length and topology changes, thus this isn't the real thing.
Nevertheless, graph effects still matter.
\cite{Stefankovic2011-hu} take a chain where the likelihoods are the likelihoods maximized across branch length.

Ranked trees~\cite{Song2006-xe}
Partitions~\cite{Gusfield2002-il}

\section{Acknowledgements}
Alex Gavruskin,
Vladimir Minin,

SAGE and GAP4 software
Alexander Hulpke

This work was funded by National Science Foundation award 1223057.

>>>>

\usepackage{amsmath,amsthm}
\usepackage{csvsimple}
\usepackage{ifthen}
\usepackage{todonotes}
%\usepackage[notref,notcite]{showkeys} % show keys for eqs, etc.
%\usepackage[right]{showlabels}
\usepackage{cite}
\usepackage{enumitem}
\usepackage{url}
\usepackage{subfigure}
\usepackage[nomarkers,nofiglist]{endfloat}
\let\MYoriglatexcaption\caption
\renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
% shift for todonotes
\usepackage[left=0.5in,right=1.5in,top=1in,bottom=1in,marginparwidth=1.25in]{geometry}
% regular margins
%\usepackage[margin=1in]{geometry}

@@

@@

Ollivier-Ricci curvature provides an intuitive formalization for quantifying the shape of a graph with respect to a random walk (or indeed any metric space, see~\cite{Ollivier2009-bw}).
Let $m_x$ and $m_y$ be probability densities of a specified random walk starting at points $x$ and $y$ of a graph $G = (V,E)$, respectively.
The Wasserstein distance~\cite{Villani2003-wv} (equivalently $L^1$ transportation, or ``earth movers distance'') between $m_x$ and $m_y$ is the minimum amount of ``work'' required to move $m_x$ to $m_y$ following the graph, that is
 $$ W_1(m_x, m_y) := \min_{\xi \in \Pi(m_x, m_y)} \sum \dspr{x,y} \xi(x,y),$$
where $\Pi(m_x, m_y)$ is the set of of assignments of probability from $m_x$ to pairs of vertices in $V \times V$.

@@

\subsection{Access time simulation}
The access time for a pair of vertices in a graph is the (random) number of iterations required to go from one of the vertices to the other in a random walk~\cite{lovasz1993random}; we were interested in the connection between curvature and access time.
In previous work, we computed mean access times (MAT) between pairs of trees in the MCMC random walks---the mean number of iterations required to move from one tree to the other.
We applied this work to demonstrate graph effects in uSPR subgraphs of real MCMC samples using MrBayes~\cite{Whidden2015-yi} using the \texttt{sprspace}~\cite{sprspace} software.

Here, to gain more insight, we used simulation to approximate the entire access time distribution.
As before, calculating these distributions for all pairs of trees would have required a tremendous amount of memory and computational power in order to get good estimates.
Here again we use the insight that the access time for a pair of trees with a simple random walk does not depend on the actual labeling of those trees, but rather only on their relative labeling.
Thus rather than enumerate access times between trees, we enumerate times between pairs of trees in a tanglegram.
For example, because the pairs
$\{(((1,2),3),4), ((1,2),(3,4))\}$ and
$\{(((1,4),3),4), ((1,4),(2,3))\}$
are the same after relabeling, their access times will have identical distributions.
Thus to calculate the empirical distributions of access times we aggregate all access times for the same tanglegram \cite{gricci, tangle}.

\begin{table}
\centering
\caption{Ordinary least squares linear multiple regression of rSPR mean access time against degree and distance.}
\subtable[5 taxa]{\csvautotabular{analysis/5-walks-200K/rspr/mean-access-5-regression.csv}}
\subtable[6 taxa]{\csvautotabular{analysis/6-walks-50K/rspr/mean-access-6-regression.csv}}
\label{tab:regressionMAT}
\end{table}

We find that the mean access time between trees $T_1$ and $T_2$ is determined by $\deg(T_1)$ and $\deg(T_2)$ (Table~\ref{tab:regressionMAT}).
Furthermore, by plotting the distribution of access times between pairs of trees with respect to their distance and curvature hints that smaller $\kappa$ slightly shifts the distribution of access times towards larger access times (Fig.~\ref{fig:short-time-kappa-access}).
We quantify this effect by defining $\delta_1$ to be the difference between the first pair of access time counts such that the second entry in the pair is nonzero.
For example, $\delta_1$ for distance 1 pairs (green lines in Fig.~\ref{fig:kappa-access}) is the count for time 1 minus the count for time 2, while $\delta_1$ for distance 3 pairs (blue lines in Fig.~\ref{fig:kappa-access}) is the count for time 2 minus the count for time 3.
Regression finds a clear influence of $\kappa$ on $\delta_1$ (Table~\ref{tab:regressionDelta}).
This confirms the intuitive interpretation of $\kappa(T_1, T_2)$ as quantifying the likelihood of a random walk going from $T_1$ to $T_2$ relatively directly, certainly before the random walk achieves stationarity.
On the other hand, if the random walk starting from $T_1$ does not quickly arrive at $T_2$ and instead achieves stationarity, the original position of the random walk is forgotten, and the access time is then a standard exponentially distributed waiting time for an event in a Poisson process (Fig.~\ref{fig:long-time-kappa-access}).

\begin{figure}
	\hspace*{\stretch{1}}
	\subfigure[\label{fig:short-time-kappa-access}]{\includegraphics[width=3in]{figs/short-time-kappa-access}}
	\hspace*{\stretch{2}}
	\subfigure[\label{fig:long-time-kappa-access}]{\includegraphics[width=3in]{figs/long-time-kappa-access}}
	\hspace*{\stretch{1}}
    \caption{\
        Distribution of rSPR $\MH$ access times for those pairs of 5-taxon trees with degree 24 that are not simple inclusions of 4-taxon pairs of trees.\
        Color signifies rSPR distance between the trees, with green, orange, and blue signifying distances of 1, 2, and 3, respectively; the saturation of the color shows curvature $\kappa(\MH; \cdot, \cdot)$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
        }
	\label{fig:kappa-access}
\end{figure}

\begin{table}
\centering
\caption{Ordinary least squares linear multiple regression of rSPR $\delta_1$ against degree, distance, and $\kappa$.}
\subtable[5 taxa]{\csvautotabular{analysis/5-walks-200K/rspr/delta1-5-regression.csv}}
\subtable[6 taxa]{\csvautotabular{analysis/6-walks-50K/rspr/delta1-6-regression.csv}}
\label{tab:regressionDelta}
\end{table}

@@

\section{Rooted SPR Neighborhoods}
Having made the connection between curvature values and access times on the SPR graphs, we now consider them theoretically.
In the absence of a bias provided by a likelihood function, curvature for the random walks we consider is entirely determined by the degree of vertices in the graph.
Thus, in order to provide bounds on curvature, we will begin by bounding differences between degrees.

@@

	We obtain the upper bounds similarly to the lower bounds, by observing that the affected at most $\frac{1}{6}$ of the mass may move through at most two fewer trees (i.e. directly between $T$ and $S$ rather than a pair of neighbors at distance $\dspr{T,S} + 2$ from each other).
	Again, this is at most one fewer tree when $\dspr{T,S}=1$.
\end{proof}

\subsection{Computational investigation and validation}
We computed all of the curvatures for pairs of trees with up to seven leaves to confirm our theoretical work with direct calculation as described before.
\todo{Update once 7 taxon runs have finished}

Indeed we see that for rSPR ... (Figure~\ref{fig:rspr-scatter}). \todo{Chris: can you fill this in?}

@@

\section{Conclusion and future work}
In summary, we have extended knowledge about the the phylogenetic rSPR graph underlying phylogenetic inference beyond previous work concerning graph diameter and vertex degree.
We done so by developing theoretical and computational frameworks to compute Ricci-Ollivier curvature of the rSPR graph.
We find that curvature, along with degree and distance, determine the early dynamics of hitting times for random walks.

We emphasize that the influence of curvature is small in this setting of equiprobable trees (Fig.~\ref{fig:short-time-kappa-access}) and that the probability of the target node in the stationary distribution is the predominant factor determining hitting times for pairs of trees (Fig.~\ref{fig:long-time-kappa-access}).
However, it is well known that MCMC takes a long time to approximate real phylogenetic posterior distributions even when the Bayesian credible set is small, and in fact our previous work showed significant graph effects in the mixing time for phylogenetic MCMC for credible sets with hundreds or thousands of trees~\cite{Whidden2015-yi}.
Thus, our next step will be to investigate curvature of MCMC with a nontrivial likelihood function, which will significantly reduce the effective size of the posterior distribution, and may lead to ``bottlenecks'' like those we have observed in real data.
We will likely follow previous work \cite{Stefankovic2011-hu} by taking a chain where tree topology likelihoods are maximized across branch length, as the classical phylogenetic likelihood \cite{felsenstein1981evolutionary} requires both branch length and topology.

In addition, we will extend this work from the rooted rSPR graph considered here to the unrooted case, which is more commonly used in phylogenetic search.
We will also explore random walks ranked trees~\cite{Song2006-xe} and models of tree space relevant for phylogenetic algorithms such as BEAST \cite{Drummond2012-ek} that infer rooted ``time-trees.''
Random walks on other discrete structures such as partitions~\cite{Gusfield2002-il} that can be expressed as certain types of trees may also form interesting subjects for future work.

\section{Acknowledgements}
The authors would like to thank Alex Gavruskin, Vladimir Minin, and Bianca Viray for helpful discussions.
They are also grateful to the authors of the SAGE and GAP4 software, especially Alexander Hulpke.
This work was funded by National Science Foundation award 1223057.
